Drought-sensitive trees survive on deep water access in a tropical forest community

### 2020-08-02 manuscript.rmd
edited a discussion paragraph

```{r erd-iso, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-iso}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE; m) versus dry season stable isotopic concentration $\\delta$^2^H (mean $\\pm$ (add error); ‰) for ", length(unique(erd.iso$sp))," canopy species from Barro Colorado Island. Data from [@Meinzer:1999]. See Table  \\@ref(tab:spcodes) for scientific names corresponding to species codes.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_phenology_Meinzer_gr.Psi.VPD.jpeg"))
```

@@

### Effective rooting depths at tree community level
\
By estimating ERD for large trees of 33 canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species species-rich tropical forests. We draw attention to the availability of data on $\delta^2^H$ on trees and water sources belowground--soil or substrate matrix water in the unsaturated vadoze zone or free water below the water table--as important for model calibration/validation. The first inverse model of ERD (water uptake depth index) developed by Chitra-Tarak et al @Chitra-Tarak:2018 lacked validation as existing methods were inadequate for extracting water bound to the soil and saprolite found in their dry forest study site (Mudumalai National Park, Southern India; Jean Riotte, pers. comm.). Our ERD model was distinctly constrained on species leaf vulnerability (PLC) curves. We benefitted from finding a strong scaling relationship between commonly collected 'soft' traits at the community scale (SLA~lamina~ and Wood Specific Gravity) and parameters of PLC curves for which we collected data on a few species (*n* = 14), a 'hard' trait. This allowed us to scale-up PLC curves at the community scale (*n* = 284 species). Whether such scaling relationships between hard and soft traits hold true for other forests and their wider applicability remains to be seen as data on PLC curves in species-rich, tropical forests are limited, but counter examples exist [see xx, yy, zz (Lawren, could you suggest references?].

\
##### Future Directions for Model Development
\
Our model estimated ERD as a single value, unlike the dynamic ERD modelled by Chitra-Tarak et al @Chitra-Tarak:2018, but future model developments should investigate seasonal dynamics of ERD, calibrated on corresponding calibration data on isotopic water sourcing depths. Our inverse ERD model was parametarised on 5-yearly growth data, with five data points over a 25 year period, which provided large degree of freedom to fit the model. Future studies could better constrain the model with higher resolution growth data such as those from dendrometers. This is especially important in modeling seasonally dynamic ERD. However, trees' stem diameter change is a sum total of not just growth, the addition of wood, but also $\pm$ changes due to hydrostatic stem flexing, whose importance increases with temporal resolution [@Chitra-Tarak.2016; @Chitra-Tarak.2014]. Use of dendrometer data would thus entail accounting for stem hydrostatic flexing [@Zweifel:2018]. In light-limited forests such as BCI, expanding ERD estimates to smaller trees, especially via parametairising over a few growth data points, would likely need model development to account for variation in trees' light environment over time and among trees' of the same size class, an issue we minimised by limiting ERD estimates to canopy trees. To estimate topographic variation in ERD within or across species a distributed hydrological model could be used. We note that data availability on soil water retention curve ($\Psi_{soil}$vs. VWC), hydraulic conductivity ($K_{soil}$) by depth and soil moisture by depth were important for effective calibration of our 1-D hydrological model, and recommend wide-spread data collection of these variables in forested sites as well as water-table levels (not collected in our site) to estimate trees' water environments. Finally, developing an inverse ERD model that accounts for movement of water within the soil-plant-atmosphere-continuum (SPAC) would be a step change from the current statistical models [this study; @Chitra-Tarak:2018]. Although data needs for parametarising such models [@Christoffersen.2016;@Sperry.1998] could be greater (e.g. retention curves for roots), limiting community-wide application, they would improve our understanding of plant physiology [@Johnson:2017].

>>>>

```{r erd-iso, eval = TRUE, echo = FALSE, message = FALSE, fig.align = 'center', out.width = "80%", fig.pos = 'H', results = "asis", fig.cap = paste0("\\label{lab-erd-iso}Modelled Effective Rooting Depth (ERD; mean $\\pm$ 1SE; m) versus dry season stable isotopic concentration $\\delta^2H$ (mean $\\pm$ (add error); ‰) for ", length(unique(erd.iso$sp))," canopy species from Barro Colorado Island. Data from [@Meinzer:1999]. See Table  \\@ref(tab:spcodes) for scientific names corresponding to species codes.")}
knitr::include_graphics(file.path("figures/PhenoDemoTraitsPsi/psi.corr_best.depth_xylem_sap_deltaD_phenology_Meinzer_gr.Psi.VPD.jpeg"))
```

@@

### Effective rooting depths at tree community level
\
By estimating ERD for large trees of 33 canopy tree species, we make an important advance in modeling effective rooting depths at the community level in species species-rich tropical forests. We note the availability of data on $\delta^2^H$ on trees and water sources belowground--soil or substrate matrix water in the unsaturated vadoze zone or free water below the water table--as important for model calibration/validation. The first inverse model of ERD (water uptake depth index, *n* = 12 species) developed by Chitra-Tarak et al @Chitra-Tarak:2018 lacked validation as existing methods were inadequate to extract water bound to the soil and saprolite found in their dry forest study site (Mudumalai National Park, Southern India; Jean Riotte, pers. comm.). Our ERD model was distinctly constrained on species leaf vulnerability (PLC) curves. We benefitted from finding a strong scaling relationship between commonly collected 'soft' traits at the community scale (SLA~lamina~ and Wood Specific Gravity) and parameters of PLC curves for which we collected data on a few species (*n* = 14), a 'hard' trait. This allowed us to scale-up PLC curves at the community scale (*n* = 284 species). Whether such scaling relationships between hard and soft traits hold true for other forests and their wider applicability remains to be seen as data on PLC curves in species-rich, tropical forests are limited, but counter examples exist [see xx, yy, zz (Lawren, could you suggest references?].

\
##### Future Directions for Model Development
\
Our model estimated ERD as a single value, unlike the dynamic ERD modelled by Chitra-Tarak et al @Chitra-Tarak:2018, but future model developments should investigate seasonal dynamics of ERD, calibrated on corresponding calibration data on isotopic water sourcing depths. Our inverse ERD model was parametarised on 5-yearly growth data, with five data points over a 25 year period, which provided large degree of freedom to fit the model. Future studies could better constrain the model with higher resolution growth data such as those from dendrometers. This is especially important in modeling seasonally dynamic ERD. However, trees' stem diameter change is a sum total of not just growth, the addition of wood, but also $\pm$ changes due to hydrostatic stem flexing, whose importance increases with temporal resolution [@Chitra-Tarak.2016; @Chitra-Tarak.2014]. Use of dendrometer data would thus entail accounting for stem hydrostatic flexing [@Zweifel:2018]. Two of our exploratory ERD models included seasonality in leaf area fraction (LeafFrac), but we did not select as the worsened the fit with growth data and ERD calibration data (Fig. S\@ref(fig:other-erd-mods)). We interpret this result as that VPD and PLC curves may have adequately explained growth, but also acknowledge that our estimate of seasonality of leaf fraction is a rough estimate based on leaf fall data and assumption of leaf presence during the wet season and we did not take into account inter-annual variation. Future studies could make use of data from new technologies such as drone based monitoring of species-specific leaf area fraction in species-rich forests [@Park.2019]. In forests such as BCI where understorey tree growth is more limited by availability of light rather than water, inversely estimating ERD for smaller tree size classes, especially from a few growth data points, may need accounting for variation in trees' light environment over time and among trees' of the same size class, an issue we minimised by limiting ERD estimates to canopy trees. To estimate topographic variation in ERD within or across species future studies may use a distributed hydrological model. We note that data availability on soil water retention curve ($\Psi_{soil}$vs. VWC), hydraulic conductivity ($K_{soil}$) by depth and soil moisture by depth were important for effective calibration of our 1-D hydrological model, and recommend wide-spread data collection of these variables as well as water-table levels (not collected in our site) in forest-inventory sites to allow for estimation of trees' water environments. Finally, developing an inverse ERD model that accounts for movement of water within the soil-plant-atmosphere-continuum (SPAC) would be a step change from the current statistical models of ERD [this study; @Chitra-Tarak:2018]. Although data needs for parametarising such models [@Christoffersen.2016;@Sperry.1998] could be greater (e.g. retention curves for roots), limiting immediate community-wide application, they promise to improve our understanding of plant physiology [e.g. see @Johnson:2017], but which may be potentially harnessed for a community level scale-up.
